\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-01-28 document/XSGRB_sample.tex
Jesper and notes on bursts

\title{XS-GRB I - The X-shooter GRB afterglow sample
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}}

@@

\abstract{In this work we present the spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed at the VLT/X-shooter up to
	\termdate. This totals spectroscopic observations of 102 individual bursts
	observed with 48 hours of the GRB trigger. Redshifts have been measured for 96
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that are unbiased with regards to intrinsic
	properties of the GRBs, the follow-up effort has been guided by the
	\textit{Swift}-telescope to produce a homogeneous sample of spectra for 91 GRBs.
	We here provide a public release of all the reduced spectra, including continuum
	estimates and telluric corrections. For completeness, we also provide reductions
	for the late-time observations of the underlying host galaxies. We provide an
	assessment of the degree of completeness with respect to the parent GRB
	population, in terms of the X-ray properties of the bursts in the sample and
	find that the sample presented here is representative of the full \textit{Swift}
	sample. For all bursts (39) for which it is possible, we provide a measurement
	of the neutral hydrogen column density, increasing the total number of published
	HI column density measurements by $\sim$ 50 per cent. %	We constrain the
	% fraction
	%	of dark bursts to be less than 28 per cent based on the detection of an
	% optical
	%	afterglow and tentatively confirm previous results that larger optical
	% darkness
	%	is correlated with increased X-ray absorption.
	This dataset provides a unique possibility to study the ISM across cosmic time,
	from the local progenitor surroundings to the galaxy wide conditions}

@@

$\gamma$-ray bursts (GRBs) provide the opportunity to derive new constraints on
a very wide range of topics in astrophysics. Examples range from small scale
phenomena relating to magnetars, properties of highly relativistic jets,
hyper/supernova explosions, the interstellar medium, dust extinction curves,
starburst galaxies, chemical and molecular abundances, escape of ionizing
radiation, the ionization state of the intergalactic medium, intervening
absoption systems to standard candles in cosmology \citep[e.g.,][]{Lyons2010,
	Molinari2007, HjorthBloom2012, Wijers1998, Prochaska2009, Savaglio2006,
	Ghirlanda2007, Vergani2009, Christensen2017}.

The \textit{Niel Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.,][]{VanParadijs2000, Ricker2004}. With the two
on-board instruments, the Burst Alert Telescope (BAT) \citep{Barthelmy2005} and
the X-Ray Telescope (XRT) \citep{Burrows2005}, \swift~is an ideal telescope for
GRB hunting. A crucial aspect of the success of the {\it Swift} mission has been
the extensive ground based follow-up observations of the afterglows and of the
host galaxies of the GRBs, involving a large community of researchers. This
fruitful collaboration has been greatly facilitated by the open data access
policy of the {\it Swift} mission. The close collaboration between detection
facilities and electromagnetic  follow-up campaigns continue to be immensely
rewarding, as recently highlighted by the simultaneous detection of
gravitational waves and light from the neutron star merger in the shape of
GW170817/GRB180817A/AT2017gfo \citep{LIGOScientificCollaboration2017a,
	LIGOScientificCollaboration2017}.

@@

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd generation
instruments at the ESO Very Large Telescope (VLT). It is a high-throughput
echelle spectrograph covering the full spectral range from the UV atmospheric
cutoff around 300 nm to the near-infrared (near-IR) K-band ending around 2500 nm
with three spectral arms (UVB, VIS, NIR). It was designed very much with
transient source follow-up in mind as the fading luminosities of such sources
makes it urgent to secure as extensive coverage as possible in the shortest
possible time. At the same time, the resolution was designed to be in the range
4000--9000 in order to be able to get a large useful spectral range between the
many sky-background emission lines in the red and near-IR spectral ranges. The
large spectral coverage allows for the spectroscopic observations of the
highest redshift GRBs.

@@

The paper is organized in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy. In Sect.~\ref{obs}, we describe the observations and the instrumental
setups, and in Sect.~\ref{proc} we detail the methodological strategies adopted
in the data reduction process and auxiliary material. In Sect.~\ref{results} we
describe the results of the survey, i.e. the efficiency of the follow-up effort
and the characteristics of the observed bursts. We also here assess the
completeness of the finally realized sample. Finally, we offer our conclusions
in Sect.~\ref{conclusions}.

@@

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item XRT started observing within 10 minutes after the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12 degrees.
	\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. The below criteria are however those
	defining the sample.}
	\item No bright, nearby stars within $ 1.8 + 0.4 \times exp[(20 - R)/2.05$] arcsec, where R is the USNO magnitude of the star)
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

@@

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to be completed. Since the GRB population is isotropically distributed
on the sky as seen from Earth and because the GRB properties do not depend on
position on the sky \citep{Meegan1992, Briggs1996}, this cut does not influence our ability
to fairly sample the underlying GRB population. The same arguments apply to
the requirement that there are no nearby bright stars. We additionally require
that the Galactic extinction is below $A_V \lesssim 0.5$ mag, based on the
extinction maps by \citet{Schlegel1998}. Because of isotropy, these additional
cuts should not influence the optical properties of the bursts themselves, only
our ability to successfully secure observations that allows us to investigate
the spectroscopic properties of GRBs.

We further discuss the effect of these selection criteria and their implication
for the completeness of the sample in Sect. \ref{completeness}.

@@

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item X-shooter mounted at the telescope
	\item GRB promptly visible
	\item Error radius < 60\arcsec.
	\item No limit on declination. 
	\item No constraint on Moon phase or distance. 
	\item Elevation in the sky > 22$^\circ$ (both at trigger time and 15 min after).
	\item Sun elevation < -12$^\circ$ (both at trigger time and 15 min after).
	\item The X-ray position must be available less than 1 hr after the GRB.
	\item The flags \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true).
\end{enumerate}

@@

There are 8 GRBs in the sample that have beem being observed with the RRM mode.
One of the RRM triggers is outside the statistical sample and two of the RRM
trigger are on short GRBs. The fastest response between \swift~GRB trigger and
the beginning of spectroscopic integration is for GRB~160410A, for which the
delay was only 8.4 minutes.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. Star symbols
	indicate bursts with measured redshifts and the circles indicate those without.
	The empty stars/circles are the positions of all the \textit{Swift} bursts
	detected until the end of 2016. A blue filling of the symbol indicates the
	position of bursts fulfilling the sample criteria specified in Sect.
	\ref{samplecrit}. The red outline is added around GRBs which have been followed
	up by X-shooter in this work. Blue symbols with red outlines represent GRBs
	that enter our sample and have X-shooter spectroscopy. The different samples
	are compared in Sect. \ref{results}. The background shows the dust maps
	presented in \citet{Schlegel1998}. Please note that we removed the background
	where the sample criteria cut ($A_V \lesssim 0.5$ mag) and replaced it with a
	white color visible at the equator. The colorbar below indicates the value of
	$A_V$ on the plot. The dotted lines indicate intervals of 30$^\circ$ in
	longitude and latitude.} \label{fig:skymap}
\end{figure*}

@@

The first GRB followed up is GRB~090313, observed on the 15th of March, 2009,
during the commissioning of X-shooter on UT3. The bursts observed during the
commissioning or science verification (GRB~090313, GRB~090530, GRB~090809,
GRB~090926) are not a part of our statistical sample. The first burst observed
after science verification and X-shooter was moved to UT2, is GRB~091018, which
thereby constitute the first burst entering our statistical sample. For all
bursts that fulfill our sample selection criteria, described in Sect.
\ref{samplecrit}, spectroscopic follow-up has been attempted with X-shooter.
Various conditions can affect our ability to follow up a given burst, and a
discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

@@

We provide an overview of all the observations in Table
\ref{tab:sample_overview} and plot the positions of all the bursts on the
celestial sphere in galactic coordinates in Fig. \ref{fig:skymap}. Away from
central zone of avoidance due to high Galactic extinction cutoff (marked in
white), the GRB positions have an isotropic distribution except in the
upper left quadrant, which cannot be observed from Paranal.

30 percent of the spectra presented here, primarily host observations, have
already been published in \citet{Kruhler2015}. We include independent
reductions of them here for completeness.

@@

In this section we describe how the final data products are produced and
subsequently post-processed. %$\sim$ 30 percent of the data released here have
% been reduced in a manner similar to what is described in \citet{Kruhler2015}.
% The two reduction schemes have been tested and are in good agreement.
All post-processing scripts developed for this dataset are made publicly
available at \url{https://github.com/jselsing/XSGRB_reduction_scripts}, along
with instructions of use.

@@

The basis for the reductions is the VLT/X-shooter pipeline, version
\texttt{2.7.1} or newer \citep{Goldoni2006, Modigliani2010}. The pipeline is
managed with the Reflex interface \citep{Freudling2013} and is used for
subtraction of bias level, flat-fielding, tracing of the echelle orders,
wavelength calibrations with the use of arc-line lamps, flux calibration using
spectrophotometric standards \citep{Vernet2010, Hamuy1994}, mirror flexure
compensation, sky-subtraction and lastly the rectification and merging
of the orders. Errors and bad pixel maps are propagated throughout the
extraction. For the initial sky-subtraction, the background has been estimated
by a running median in regions adjacent to the object trace clear of
contaminating sources. Due to the broken ADC, for some objects there is
curvature in the object trace along the dispersion axis of the slit. This means
that for some bursts, the initial sky-estimate has been made from a limited
number of pixels in the spatial direction. By doing a subtraction of
the sky background on the un-rectified image we ensure that the bulk of the sky
background is not redistributed by the rectification process.

The image is rectified onto an equidistant grid with a dispersion sampling of
0.02 nm/pixel and a 0.16 "/pixel spatial sampling for the UVB and VIS arm and
0.06 nm/pixel with a 0.21 "/pixel in the NIR arm.  Because the tabulated
resolution is a lower limit, we choose a sampling of 0.02 nm/pixel to ensure
that the lowest wavelength part of neither of the arms have a sampling lower
than the Nyquist sampling rate of 2 pixels per resolution FWHM.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as a single observation and then subsequently combined to form a
single image. We employ this strategy so that we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images where the noise in each pixel is dominated by
Poisson noise it is important to estimate the background variance in a large
enough region, so that any correlation between the signal and the weights are
removed. To this end, the weight map is generated by a running median window
over the variance map produced by the pipeline, where the trace has been masked
and width of the window is chosen to be wide enough for median variance to be
generated on the basis of several hundred pixels. This weighting scheme
automatically also optimally combines images of different exposure times or
images where the background is varying, which is often the case when a burst has
been observed close to twilight

@@

Using this method for all observations we additionally get a spectrum of the
sky, which we can use to calibrate the wavelength solution.

@@

Since X-shooter is installed at the VLT Cassegrain focus it is prone to
flexures. The flexures modify the projection of the slit on the detector with
respect to the one obtained in daytime calibration. This requires a modification
of the wavelength solution in order to correctly process the night-time data.
Part of this correction is performed by the pipeline using the frames taken
during the X-shooter Active Flexure Compensation procedure \footnote{X-shooter
	User Manual available at
	https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}. The
remaining offset is corrected by cross-correlating the observed sky spectrum
with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum,
estimated as the mode of all flux values, has been subtracted. To get the
correct seeing PSF with which to convolve the synthetic sky an initial
refinement of the wavelength solution has been obtained by cross-correlating
the observed sky with an unconvolved synthetic sky. This preliminary wavelength
calibration is applied to the observed sky. The synthetic spectrum is then
convolved with an increasing seeing PSF and the width that minimizes $\chi^2$
with the updated observed sky is chosen to be the effective sky-PSF. Using the
synthetic sky with the matched resolution, a final wavelength calibration can
then be calculated by cross-correlating the observed sky with the correctly
broadened sky spectrum, as a function of a velocity offset. Both a
multiplicative and an additive offset to the wavelength calibration has been
tested, but in terms of $\chi^2$, the model with only a multiplicative offset is
preferred. The resulting offsets, which were smaller than 0.01 nm in the UVB and
VIS data and smaller than 0.05 nm in the NIR spectra, but changing over short
period of time were applied to the corresponding spectra. Using the convolved
synthetic sky, the pixels containing the brightest sky lines have been added to
the bad pixel map.

@@

To get the shape and the position of the spectral PSF as a function of location
on the image, we need to chose a model which can represent how the light falls
on the slit. We know from \citet{Trujillo2001} that a Moffat function
\citep{Moffat1969} with an index of $\beta = 4.765$ adequately describes an
imaging PSF due to atmospheric turbulence, but because of aberrations in the
dispersion elements and the rectification process, the PSF we are trying to
model is different from this profile. To allow for flexibility in the model, we
have chosen the Voigt function as a model for the spectral PSF and we describe
how this is evaluated in App. \ref{voigt}. Since additionally, the host galaxy
could also have a contribution to the image profile, this choice allows for the
required freedom if additional flux is in the wings of the profile.

To guide the guess position of the trace on the slit as a function of
wavelength, we have used the analytic prescription for the trace position
described in \citet{Filippenko1982}, where the header keywords of the
observations have been queried for the ambient conditions which controls the
degree to which the trace changes position in the spatial direction.

@@

The spectra are corrected for Galactic extinction using the $E(B-V)$ value from
the dust maps of \citet{Schlegel1998} with the update in
\citet{Schlafly2011}\footnote{Queried from
	http://irsa.ipac.caltech.edu/applications/DUST/index.html using
	\citet{astroquery}}, and the extinction curve by \cite{Cardelli1989} with a
total to selective extinction $R_V = 3.1$. The wavelengths of the extracted
1D-spectra are moved to vacuum, corrected for barycentric motion, and the
wavelength recalibration described in Sect. \ref{wavecal} is applied. Pixels
with pixel-to-pixel variation larger than $50 \sigma$ are additionally added to
the bad pixel map.

@@

The highest priority method is using the GRB afterglow continuum itself, where
the atmospheric conditions have directly been imprinted on the spectrum. The
telluric features can directly be fit with an atmospheric model
(\texttt{Molecfit} \citep{Smette2015,
	Kausch2015}\footnote{\url{http://www.eso.org/sci/software/pipelines/skytools/molecfit}}), which can then be used to correct for the absorption. The accuracy of the correction depends on the S/N per pixel for the target spectrum, where we have chosen the requirement that the afterglow continuum spectrum has a median signal-to-noise higher than a value of 10.

If the afterglow is not sufficiently bright, telluric standard stars observed
close in the time to the GRB can be used as a proxy for the atmospheric
condition during the GRB observation. Here we employ the telluric correction
method that has been developed in \citet{Selsing2015}, where a library of
synthetic templates is fit to the observed telluric standard.

In the last case, where the object is neither bright enough, or there for some
reason have not been observed a telluric standard, we rely on the synthetic sky
model by \citet{Noll2012} and \citet{Jones2013} for which we generate a
synthetic transmission spectrum,  where the ambient parameters for the
observations have been used.

>>>>

\title{XS-GRB I - The X-shooter GRB afterglow sample
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program IDs: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260.}}

@@

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with VLT/X-shooter up to
	\termdate. This totals spectroscopic observations of 102 individual bursts
	observed with 48 hours of the GRB trigger. Redshifts have been measured for 96
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that are unbiased with regards to intrinsic
	properties of the GRBs, the follow-up effort has been guided by the
	\textit{Swift}-telescope to produce a homogeneous sample of spectra for 91 GRBs.
	We here provide a public release of all the reduced spectra, including continuum
	estimates and telluric corrections. For completeness, we also provide reductions
	for the late-time observations of the underlying host galaxies. We provide an
	assessment of the degree of completeness with respect to the parent GRB
	population, in terms of the X-ray properties of the bursts in the sample and
	find that the sample presented here is representative of the full \textit{Swift}
	sample. For all bursts (41) for which it is possible, we provide a measurement
	of the neutral hydrogen column density, increasing the total number of published
	HI column density measurements by $\sim$ 50 per cent. %	We constrain the
	% fraction
	%	of dark bursts to be less than 28 per cent based on the detection of an
	% optical
	%	afterglow and tentatively confirm previous results that larger optical
	% darkness
	%	is correlated with increased X-ray absorption.
	This dataset provides a unique possibility to study the ISM across cosmic time,
	from the local progenitor surroundings to the galaxy wide conditions}

@@

$\gamma$-ray bursts (GRBs) provide the opportunity to derive constraints on
a very wide range of topics in astrophysics. Examples range from small scale
phenomena relating to magnetars, properties of highly relativistic jets,
hyper/supernova explosions, the interstellar medium, dust extinction curves,
starburst galaxies, chemical and molecular abundances, escape of ionizing
radiation, the ionization state of the intergalactic medium, intervening
absoption systems to standard candles in cosmology \citep[e.g.,][]{Lyons2010,
	Molinari2007, HjorthBloom2012, Wijers1998, Prochaska2009, Savaglio2006,
	Ghirlanda2007, Vergani2009, Christensen2017}.

The \textit{Niel Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of the
intergalactic medium, which was already hinted at by results from earlier
missions \citep[e.g.,][]{Paradijs2000, Ricker2004}. With the two on-board
instruments, the Burst Alert Telescope (BAT) \citep{Barthelmy2005} and the X-Ray
Telescope (XRT) \citep{Burrows2005}, \swift~is an ideal telescope for GRB
hunting. A crucial aspect of the success of the {\it Swift} mission has been the
extensive ground based follow-up observations of the afterglows and of the host
galaxies of the GRBs, involving a large community of researchers. This fruitful
collaboration has been facilitated by the open data access policy of the
{\it Swift} mission. The close collaboration between detection facilities and
electromagnetic  follow-up campaigns continue to be immensely rewarding, as
recently highlighted by the simultaneous detection of gravitational waves and
light from the neutron star merger in the shape of GW170817/GRB180817A/AT2017gfo
\citep{LIGOScientificCollaboration2017a, LIGOScientificCollaboration2017}.

@@

The X-shooter spectrograph \citep{Vernet2011} was the first of the 2nd
generation instruments at the ESO Very Large Telescope (VLT). It was designed
very much with transient source follow-up in mind as the fading luminosities of
such sources makes it urgent to secure as extensive coverage as possible in the
shortest possible time. At the same time, the resolution was designed to be in
the range 4000--9000 in order to be able to get a large useful spectral range
between the many sky-background emission lines in the red and near-IR spectral
ranges. The large spectral coverage allows for the spectroscopic observations of
the highest redshift GRBs.

@@

The paper is organized in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy. In Sect.~\ref{obs}, we describe the observations and the instrumental
setups, and in Sect.~\ref{proc} we detail the methodological strategies adopted
in the data reduction process and auxiliary material. In Sect.~\ref{results} we
describe the results of the survey, i.e. the efficiency of the follow-up effort
and the characteristics of the observed bursts. We also here assess the
completeness of the finally realized sample. Finally, we offer our conclusions
in Sect.~\ref{conclusions}. We use the $\Lambda$CDM cosmology parameters
provided by \citet{Planck2015} in which the universe is flat with $H_0 =
67.7$\,km\,s$^{-1}$ Mpc$^{-1}$ and $\Omega_\mathrm{m} = 0.307$.

@@

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item XRT started observing within 10 minutes after the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes, 30 degrees above the horizon, with the Sun below $-12$ degrees\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. These criteria are, however, those
	defining the sample.}.
	\item No bright, nearby stars within $ 1.8 + 0.4 \times exp[(20 - R)/2.05$] arcsec, where R is the USNO magnitude of the star)
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

@@

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to be completed. Since the GRB population is isotropically
distributed on the sky as seen from Earth and because the GRB properties do not
depend on position on the sky \citep{Meegan1992, Briggs1996, Ukwatta2016}, this
cut does not influence our ability to fairly sample the underlying GRB
population. The same arguments apply to the requirement that there are no nearby
bright stars. We additionally require that the Galactic extinction is below $A_V
\lesssim 0.5$ mag, based on the extinction maps by \citet{Schlegel1998,
	Schlafly2011}. Due to the isotropic distribution of GRBs on the sky, these
additional cuts should not influence the optical properties of the bursts
themselves, only our ability to successfully secure the observations that allows
us to investigate the spectroscopic properties of GRBs.

Bursts that fulfill these selection criteria are what we define as our
\textit{statistical} sample. Using this sample we will be able to address
population properties of \swift-detected bursts. We further discuss the effect
of these selection criteria and their implication for the completeness of the
sample in Sect. \ref{completeness}.

@@

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item X-shooter mounted at the telescope
	\item GRB promptly visible
	\item Error radius < 60\arcsec.
%	\item No limit on declination. 
%	\item No constraint on Moon phase or distance. 
	\item Elevation in the sky $> 22^\circ$ (both at trigger time and 15 min after).
	\item Sun elevation $< -12^\circ$ (both at trigger time and 15 min after).
	\item The X-ray position must be available less than 1 hr after the GRB.
	\item The flags \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true).
\end{enumerate}

@@

There are 8 GRBs in the sample that have beem being observed with the RRM mode.
One of the RRM triggers is outside the sample we use to address the statistical
properties of GRB afterglows and two of the RRM trigger are on short GRBs. The
fastest response between \swift~GRB trigger and the beginning of spectroscopic
integration is for GRB~160410A, for which the delay was only 8.4 minutes.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. Star symbols
	indicate bursts with measured redshifts and the circles indicate those without.
	The empty stars/circles are the positions of all the 1266 \textit{Swift} bursts
	detected until \termdate. A blue filling of the symbol indicates the
	position of bursts fulfilling the sample criteria specified in Sect.
	\ref{samplecrit}. The red outline is added around GRBs which have been followed
	up by X-shooter in this work. Blue symbols with red outlines represent GRBs
	that enter our sample and have X-shooter spectroscopy. The different samples
	are compared in Sect. \ref{results}. The background shows the dust maps
	presented in \citet{Schlegel1998}. Please note that we removed the background
	where the sample criteria is violated ($A_V > 0.5$ mag) and replaced it with a
	white color visible at the equator. The colorbar below indicates the value of
	$A_V$ on the plot. The dotted lines indicate intervals of 30$^\circ$ in
	longitude and latitude.} \label{fig:skymap}
\end{figure*}

@@

The first GRB followed up is GRB~090313, observed on the 15th of March, 2009,
during the commissioning of X-shooter on UT3. The bursts observed during the
commissioning or science verification (GRB~090313, GRB~090530, GRB~090809,
GRB~090926) are not a part of the sample we use to investigate the statistical
properties of GRB afterglows. The first burst observed after science
verification and X-shooter was moved to UT2, is GRB~091018, which is the first
burst entering our statistically homogeneous sample. For all bursts that fulfill
our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up has been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of these
conditions and their consequences for the sample is included in Sect.
\ref{badbursts}.

@@

We provide an overview of all the observations in Table
\ref{tab:sample_overview} and plot the positions of all the bursts on the
celestial sphere in galactic coordinates in Fig. \ref{fig:skymap}. Away from the
central zone of avoidance due to high Galactic extinction cutoff (marked in
white), the GRB positions have an isotropic distribution, except in the upper
left quadrant which cannot be observed from Paranal.

30 percent of the spectra presented here, primarily host observations, have
already been published in \citet{Kruhler2015}. Single bursts have additionally
been published, based on unusual properties in their afterglows. We present
individual notes on all the X-shooter spectra and their previous use in App.
\ref{notes}. We include independent reductions of them here for completeness.

@@

In this section we describe how the final data products are produced and
subsequently post-processed. %$\sim$ 30 percent of the data released here have
% been reduced in a manner similar to what is described in \citet{Kruhler2015}.
% The two reduction schemes have been tested and are in good agreement.
All post-processing scripts developed for this dataset are made publicly
available at \url{https://github.com/jselsing/XSGRB_reduction_scripts}.

@@

The basis for the reductions is the VLT/X-shooter pipeline, version
\texttt{2.7.1} or newer \citep{Goldoni2006, Modigliani2010}. The pipeline is
managed with the Reflex interface \citep{Freudling2013} and is used for
subtraction of bias level, flat-fielding, tracing of the echelle orders,
wavelength calibrations with the use of arc-line lamps, flux calibration using
spectrophotometric standards \citep{Vernet2010, Hamuy1994}, mirror flexure
compensation, sky-subtraction and lastly the rectification and merging of the
orders. Errors and bad pixel maps are propagated throughout the extraction. For
the initial sky-subtraction, the background has been estimated by a running
median in regions adjacent to the object trace clear of contaminating sources.
Due to the broken ADC, for some objects there is curvature in the object trace
along the dispersion axis of the slit. This means that for some bursts, the
initial sky-estimate has been made from a limited number of pixels in the
spatial direction. The subtraction of the sky background on the un-rectified
image ensures that the bulk of the sky background is not redistributed by the
rectification process.

The image is rectified onto an equidistant grid with a dispersion sampling of
0.02 nm/pixel and a 0\farcs16 per pixel spatial sampling for the UVB and VIS
arm and 0.06 nm/pixel with a 0\farcs21 per pixel in the NIR arm.  Because
the tabulated resolution is a lower limit, we choose a sampling of 0.02 nm/pixel
to ensure that the lowest wavelength part of neither of the arms have a sampling
lower than the Nyquist sampling rate of 2 pixels per resolution FWHM.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as a single observation and then subsequently combined to form a
single image. We employ this strategy so that we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images where the noise in each pixel is dominated by
Poisson noise, it is important to estimate the background variance in a large
enough region, so that any correlation between the signal and the weights are
removed. To this end, the weight map is generated by a running median window
over the variance map produced by the pipeline, where the trace has been masked
and the width of the window is chosen to be wide enough for median variance to
be generated on the basis of several hundred pixels. This weighting scheme
automatically also optimally combines images of different exposure times or
images where the background is varying, which is often the case when a burst has
been observed close to twilight

@@

Using this method for all observations we additionally get a spectrum of the
sky, which we can use to recalibrate the wavelength solution.

@@

Since X-shooter is installed at the VLT Cassegrain focus it is prone to
flexures. The flexures modify the projection of the slit on the detector with
respect to the one obtained in daytime calibration. This requires a modification
of the wavelength solution in order to correctly process the night-time data.
Part of this correction is performed by the pipeline using the frames taken
during the X-shooter Active Flexure Compensation procedure\footnote{X-shooter
	User Manual available at
	https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}. The
remaining offset is corrected by cross-correlating the observed sky spectrum
with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum,
estimated as the mode of all flux values, has been subtracted. To get the
correct seeing PSF with which to convolve the synthetic sky an initial
refinement of the wavelength solution has been obtained by cross-correlating the
observed sky with an unconvolved synthetic sky. This preliminary wavelength
calibration is applied to the observed sky. The synthetic spectrum is then
convolved with an increasing seeing PSF and the width that minimizes $\chi^2$
with the updated observed sky is chosen to be the effective sky-PSF. Using the
synthetic sky with the matched resolution, a final wavelength calibration can
then be calculated by cross-correlating the observed sky with the correctly
broadened sky spectrum, as a function of a velocity offset. Both a
multiplicative and an additive offset to the wavelength calibration has been
tested, but in terms of $\chi^2$, the model with only a multiplicative offset is
preferred. The resulting offsets, which were smaller than 0.01 nm in the UVB and
VIS data and smaller than 0.05 nm in the NIR spectra, but changing over short
periods of time were applied to the corresponding spectra. Using the convolved
synthetic sky, the pixels containing the brightest sky lines have been added to
the bad pixel map.

@@

To get the shape and the position of the spectral PSF as a function of location
on the image, we need to chose a model which can represent how the light falls
on the slit. We know from \citet{Trujillo2001} that a Moffat function
\citep{Moffat1969} with an index of $\beta = 4.765$ adequately describes an
imaging PSF due to atmospheric turbulence, but because of aberrations in the
dispersion elements and the rectification process, the PSF we are trying to
model is different from this profile. To allow for flexibility in the model, we
have chosen the Voigt function as a model for the spectral PSF and we describe
how this is evaluated in App. \ref{voigt}. Since the host galaxy could also give
a contribution to the image profile, this choice allows for the required freedom
if additional flux is in the wings of the profile.

To guide the first guess position of the trace on the slit as a function of
wavelength, we have used the analytic prescription for the trace position
described in \citet{Filippenko1982}, where the header keywords of the
observations have been queried for the ambient conditions which controls the
degree to which the trace changes position in the spatial direction.

@@

The spectra are corrected for Galactic extinction using the $E(B-V)$ value from
the dust maps of \citet{Schlegel1998} with the update in
\citet{Schlafly2011}\footnote{Queried from
	http://irsa.ipac.caltech.edu/applications/DUST/index.html using \texttt{astroquery}
	\citep{astroquery}.}, and the extinction curve by \cite{Cardelli1989} with a
total to selective extinction $R_V = 3.1$. The wavelengths of the extracted
1D-spectra are moved to vacuum, corrected for barycentric motion, and the
wavelength recalibration described in Sect. \ref{wavecal} is applied. Pixels
with pixel-to-pixel variation larger than $50 \sigma$ are additionally added to
the bad pixel map.

@@

The highest priority method is using the GRB afterglow continuum itself, where
the atmospheric conditions have directly been imprinted on the spectrum. The
telluric features can directly be fit with an atmospheric model
(\texttt{Molecfit} \citep{Smette2015,
	Kausch2015}\footnote{\url{http://www.eso.org/sci/software/pipelines/skytools/molecfit}}), which can then be used to correct for the absorption. The accuracy of the correction depends on the S/N per pixel of the target spectrum, where we have chosen the requirement that the afterglow continuum spectrum has a median signal-to-noise higher than a value of 10.

If the afterglow is not sufficiently bright, telluric standard stars observed
close in time to the GRB can be used as a proxy for the atmospheric
condition during the GRB observation. Here we employ the telluric correction
method that has been developed in \citet{Selsing2015}, where a library of
synthetic templates is fit to the observed telluric standard.

In the last case, where the object is neither bright enough, or there for some
reason have not been observed a telluric standard, we rely on a synthetic sky
model \citep{Noll2012, Jones2013} for which we generate a
synthetic transmission spectrum, where the ambient parameters for the
observations have been used.

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$,
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$r$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption and emission, including
	absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
	$\sim$ 4000 $\AA$ is due to \lya~in the host. We have marked the most prominent
	lines seen in GRB afterglows from \citet{Christensen2011a}. The regions of most
	severe telluric absorption is highlighted by a grey-shading the background.
	Additionally, three intervening systems are seen in this sightline. This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
\label{fig:spectrum}
\end{figure*}
